Revealing Benefits, Costs, and Tradeoffs of Spatial Management in Marine Systems Arrising from Coupled Economic and Ecological Dynamics

### 2023-03-28 03_marlin_paper.qmd
passing back to tuna for edits
passing back to abalone for a final top to bottom run

# simple function to make prettier labels
titler <- function(x){

  stringr::str_to_title(stringr::str_replace_all(x,"_"," "))

@@

Abstract: Designing effective spatial management strategies is
challenging because marine ecosystems are highly dynamic and opaque, and
extractive entities such as fishing fleets respond endogenously to
ecosystem changes in ways that depend upon ecological and policy
context. We present a tool -- marlin -- to efficiently simulate the
bio-economic dynamics of marine systems, designed to support both
management and research. We demonstrate `marlin’s` capabilities by
focusing on two case studies on the conservation and food production
impacts of marine protected areas (MPAs): a coastal coral reef and a
pelagic tuna fishery. In the coastal coral reef example, we explore how
heterogeneity in species distributions and fleet preferences can affect
distributional outcomes of MPAs. In the pelagic case study, we
demonstrate how our model can be used to assess the climate resilience
of different MPA design strategies, as well as the climate sensitivity
of different fishing fleets. This paper demonstrates how intermediate
complexity simulation of coupled bio-economic dynamics can help
communities predict and potentially manage trade-offs between
conservation, fisheries yields, and distributional outcomes resulting
from spatial management.

@@

Communities around the world are increasingly looking to spatial
strategies for managing marine ecosystems . For example, the "30x30"
movement calls for implementation of marine protected areas (MPAs), a
form of spatial management, across 30% of the world's oceans by the year
2030 [@grorud-colvert2021 and references therein]. However , designing
spatial management strategies to achieve desired objectives -- which may
include recovery of overfished species, increased food production and
economic well being, and equitable distribution of benefits - is not
straightforward. Different species have different resilience to fishing,
distributions in space and time, and value to fishing fleets; fishing
fleets themselves can have varying ranges of incentives and reactions to
policy structures. These dynamics pose challenges even for
single-species assessment and management strategies [@field2006a], which
are only amplified when we consider management policies specifically
designed around spatial features in the oceans.

For example, policies such as MPAs designed for both conservation and
food production must consider factors such as the optimal size and
placement of a protected area network given a wide array of life
history, species distributions, exploitation levels, fleet dynamics, and
policy-dependent behavior [@reimer2017], all of which may change over
time, particularly given the impacts of climate change. Efforts to
effectively design spatial management strategies such as MPAs are
further constrained by a lack of empirical evidence describing the size
[@ovando2021a] and time required [@nickols2019] for MPAs to produce
substantial effects across a range of social-ecological settings
[@mcclanahan2021].

Many models have been developed to support the theory and design of
spatial management strategies (see @fulton2015 for a review of existing
methods). However, they have tended to be either highly complex and
tactical models designed for use in a specific location, or to be
extremely stylized representations intended to provide theoretical
insights with less relevance for specific applications. This lack of
accessible models suited for more tactical application presents a
challenge to stakeholders charged with marine spatial planning
exercises. It also presents a barrier to the scientific community,
wherein comparisons across spatial management simulation studies are
clouded by discrepancies in underlying model structures beyond
differences in the core phenomenon in question.

To help address this challenge, we present a novel bio-economic modeling
tool called [`marlin`](https://danovando.github.io/marlin/) that allows
for efficient simulation of spatio-temporal biological and economic
dynamics. `marlin` allows users to simulate the impacts of marine
management policies across a range of species targeted by various fleets
across heterogeneous and dynamic seascapes. This model can help users
seek Pareto-optimal solutions that produce win-wins or minimize
tradeoffs across multiple management objectives, such as food production
and conservation [@lester2013].

Here we present the core methods and functionality of `marlin`, and
demonstrate its use in two case-study applications; a coral reef system
and a pelagic system. Our results show how interacting ecological,
economic, and design attributes can affect the degree of benefits,
costs, and tradeoffs between conservation and food production outcomes
of MPAs. More broadly, this paper both demonstrates the functionality of
our model and the critical need for considering sufficiently realistic
coupled ecological and economic dynamics in policy evaluations.

@@

The front end of the software package accompanying this paper,
[`marlin`](https://danovando.github.io/marlin/), is written in R
[@rcoreteam2021] to facilitate use, while the underlying population
model is written in C++, integrated through the `Rcpp` package
[@eddelbuettel2018], for speed. Users on standard computers should be
able to simulate one age-structured population distributed across a 100
patch system for 20 years in fractions of a second.

@@

We define a few commonly used terms here. *Yields* refers to the volume
of catch from fishing activity. *Spawning stock biomass* refers to the
total biomass of reproductively mature fish in the water, a function of
numbers, weight, and sexual maturity at age. We measure the size of the
population by th ratio of the *spawning stock biomass* (SSB) in a given
time step relative to *unfished spawning stock biomass* (SSB0),
SSB/SSB0. So, a SSB/SSB0 value of 1 means that the population is
unfished, a SSB/SSB0 value of 0 means the population is extinct.

@@

These species can be caught by fishing fleets. A fishing fleet is
defined in `marlin` by a set of fishers that have the same fishing
skill, prices, and selectivity for individual species (each of which we
denote as a *metier*). For example, both a longline and purse seine
fleet may capture bigeye (*Thunnus obesus*) and skipjack tunas, but the
longline fleet can be made much more likely to catch bigeye than
skipjack, and *vice versa*. Each fishing fleet distributes its fishing
effort in space according to a specified spatial allocation rule (see
*Spatial Allocation of Effort* ), for example in proportion to
profit-per-unit-effort, conditional on management policies in places
such as quotas and/or the presence of any spatial restrictions such as
MPAs.

@@

`marlin` also allows users to specify closed fishing seasons for one or
more species in the system. When all species in a simulation share the
same closed seasons, no fishing occurs in those seasons. When not all
species share the same closed seasons, fishing still occurs in the
closed seasons, but the fleet avoids catching the species for which
fishing is closed in a given season, assuming no discard mortality. When
closed seasons are in effect, the amount of effort in the next open
season is a function of the effort and profits in the last open fishing
season. Lastly, managers can specify locations of no-take MPAs, which
could in theory change in size and location. When MPAs are implemented,
fishing effort that used to operate inside the MPAs can either leave the
fishery, or be redistributed outside the MPA (the default behavior).

@@

Each fishing fleet decides how to allocate its effort in space based on
one of four possible spatial allocation strategies. The ideal free
distribution (IFD) is the standard method for distributing fishing
fleets in spaces (see @gillis2003 and references therein). However,
analytical solutions to the IFD present a number of complications for
our model. Primarily, our model is capable of tracking multiple fleets
at once. This means that the IFD for a single fleet would commonly be
modeled as a Nash-equilibrium based on the actions of each of the
individual fleet conditional on the actions of all other fleets. While
possible to implement, this would slow down our model runs to the point
of practically preventing large-scale evaluation of the spatial
management policies.

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*advection* (follow currents), and *taxis* (active movement towards
preferred habitat) and *diffusion* (essentially remaining variation in
movement not explained by advection or taxis). For now, `marlin` focuses
just on the diffusion and taxis components of this model, assuming that
advection is zero, though future extensions could easily incorporate
advection vectors from oceanographic models. In this way, `marlin`
allows users to run anything from a simple Gaussian dispersal kernel up
to a system governed by species that passively diffuse out from a core
habitat defined by a dynamic thermal range.

We provide a brief overview of the the general CTMC method here, see
@thorson2021a for a detailed description. Under this framework, movement
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of the sum of the diffusion $\boldsymbol{D}$ and taxis
$\boldsymbol{\tau}$ matrices scaled by the width of the time step
$\Delta_{t}$ and the length of the edge of each patch $\Delta_d$ that
the model is running under. So for example, if we are running a model on
a monthly timestep, and we define the units of time in years, then
$\Delta_{t} = 1/12$. Similarly, if we are running a model where one
patch equals 100km^2^, then $\Delta_d = 100km$. This parameterization
allows users to set the effective area of the spatial domain through two
avenues; the number of patches, which effectively scales the resolution
of the model, and the area of each patch, which scales the spatial
extent of the simulation.

@@

The individual components of the diffusion matrix are filled based on an
adjacency matrix, which defines whether two patches are both adjacent
and water (as opposed to land or another physical barrier), and a
diffusion rate $D$ defined in units of length of the side of a patch per
unit of time. For example, if we are defining the time units as years
and the distance units as kilometers, for a tuna $D$ might be 1,000
$\frac{KM}{Year}$. The $\Delta_{t}$ and $\Delta_{d}$ parameters then
translate the movement rate $D$ to match the time step and patch size
used in a simulation. This "scale free" parameterization means that
appropriate value of $D$ can be identified for a species and then set,
regardless of the time step or patch size used in the simulation model
itself. These parameters are then used to calculate the value of a
particular entry the diffusion matrix (*d*).

$$ 
d_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}^2}D & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} d_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
\end{cases}
$$ {#eq-diffusion}

The taxis matrix $\boldsymbol{\tau}$ is built off of the difference in
log habitat quality between each patch (the habitat gradient), the
adjacency matrix, and temporal and spatial resolution of the model
($\Delta_{t}$ and $\Delta_{p}$), as well as a scalar term $\lambda$.
$\lambda$ scales the magnitude of the habitat gradient between patches
relative to the diffusion parameter $D$. When the diffusion and taxis
matrices are estimated jointly from actual data in the manner of
@thorson2021a, $\lambda$ should be set to zero. However, when users are
manually creating both the diffusion and taxis matrices for simulation,
the relative strength of diffusion to taxis must be defined by the user.
Setting higher values of $\lambda$ allows users to simulate a scenario
where taxis is much more significant than diffusion.

$$
\tau_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}}(1 + \lambda d_{s,f})e^{log(h(p2,t,s,a) - log(h(p1,t,s,a))} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} \tau_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
  \end{cases}
$$ {#eq-taxis}

@@

While this CTMC approach to movement simulation is to date not commonly
seen in the marine modeling literature, it has numerous advantages that
warrant its seeming complexity. First, when only diffusion is present,
the model will generalize to the familiar dynamics of a Gaussian
dispersal kernel. However, the time component *t* in @eq-movement allows
the movement model to be run on arbitrarily fine time steps. Second, the
taxis model allow for more active habitat choices by species, allowing
us to simulate preferences of species in space and time efficiently.
Lastly, this form has the advantage that the parameters are directly
estimable from real data. So, if provided with for example spatial
abundance data and a tagging study, users can estimate the diffusion and
taxis movement parameters in the same manner as @thorson2021a, and then
pass their estimated parameters to `marlin` for simulation.

@@

Recruitment (i.e. the number of age 1 individuals entering the
population) follows Beverton-Holt dynamics parameterized around
steepness (*h*) with log-normally distributed recruitment deviates
$\epsilon$. When steepness is one recruitment is independent of spawning
biomass. As steepness approaches 0.2 recruitment becomes a linear
function of spawning biomass. We allow for five variants in the timing
of density dependent recruitment:

@@

4.  Post-dispersal density dependence: Larvae are distributed throughout
    the system, and then density dependent recruitment occurs based on
    the density of adult biomass at the destination patch.

@@

Fisheries management is often concerned with measuring stock status
relative to maximum sustainable yield (MSY) based reference points,
though the exact level of stock status desired by management or societal
objectives may vary widely. MSY based reference points present a problem
for a multi-fleet and spatial-temporal model such as `marlin`. MSY and
the fishing mortality rate that would produce MSY, $F_{MSY}$, are a
function of fishery selectivity. Fishery selectivity in this model can
vary by fleet, and species can be distributed unevenly in both space and
time. This means that the net effective fishing selectivity on a species
can vary depending on the dynamics at a given moment, making the
definition of an equilibrium concept such as MSY challenging.

@@

Each of the case studies contains too many parameters and options to be
succinctly presented in the text here, and we are concerned that
attempting to duplicate all of the parameters of the model here would
simply introduce error. As such, readers should consult the accompanying
code to view the precise details of each simulation. Targeted tactical
applications must carefully consider and document all decisions made
around model parameters.

@@

Across all of our case studies we make use of two potential "rule of
thumb" MPA design strategies:

@@

To our knowledge, MPAs are never randomly placed; their locations are
decided based on stakeholder negotiation given available information and
objectives. In theory, the design of MPA networks can be optimized
through the use of a modeling framework, and depending on the validity
of the model, this process is likely to produce better outcomes than
manually-designed strategies [@rassweiler2012; @rassweiler2014].
However, designing optimal MPA networks becomes increasingly difficult
as the range of objectives and the complexity of the model increase.
Therefore, while `marlin` provides some support for
[optimization](https://danovando.github.io/marlin/articles/optimize-mpas.html),
we focus here on the design and performance of these more "rule of
thumb" design strategies that may be more accessible to a wider range of
users. We allow all MPAs to be designed in a "mosaic" fashion, but users
can easily extend the analysis to compare outcomes between "contiguous"
(MPA is made up of one continuously connected block) and "mosaic" (MPAs
can be separated in space) MPA designs [@pons2022] by developing and
passing their own MPA design algorithms to the model.

#### Coral Reefs

In our coral reef example, we model the dynamics of four tropical
Pacific species: a grouper (*Epinephelus fuscoguttatus*), a shallow-reef
snapper (*Lutjanus malabaricus*), a deep-water snapper (*Pristipomoides
filamentosus*), and a reef shark (*Carcharhinus amblyrhynchos*). The
simulated groupers undergo a mass migration to a spawning aggregation
once per year, followed by the sharks. Shallow-reef snappers stay in
reefs closer to shore year-round, while deep-water snappers stay in the
deeper reefs (@fig-coral).

These species are targeted by two different fleets. Fleet One primarily
targets the grouper and nearshore snapper populations, but will land any
incidentally captured sharks. Fleet One has a dome shaped selectivity
pattern for both grouper and shallow-reef snapper due to demand for
"plate sized" fish [@kindsvater2017], and logistic selectivity for
sharks, as any shark caught is retained. Fleet One is totally dependent
on fishing for their livelihood, meaning the local community takes
advantage of every possible opportunity to fish, and as such we model it
as a "constant effort" fishery.

Fleet Two primarily targets the deep-water snapper population. While
plate-sized snapper are still prized from this population, we model
Fleet Two's selectivity as logistic due to high levels of discard
mortality for deep-water snapper resulting from barotrauma. Fleet Two
catches some grouper, though less than Fleet One, and receives no price
for sharks due to the requirements of a certification program through
which they sell their deep-water snapper. Accidental captures (bycatch)
do occur, which results in mortality. Fleet Two operates under
open-access dynamics, as fishing is not the primary means of
subsistence; short-term effort expands and contracts in response to
profitability of the primarily grouper-driven fishery.

Each fleet has a different home-port location, and incurs an additional
cost for fishing each patch in the system in proportion to the Euclidean
distance from each patch to each fleet's home port.

We use `marlin` to simulate the outcomes for both food production and
conservation for each of the species and both of the fleets as a
function of both MPA size and MPA design strategy. For this exercise,
MPAs are placed with perfect information and have no design constraints
for continuity (i.e. the selected MPA design can be as patchwork as
dictated by the data). We ran the simulation on a quarterly level
($\Delta_t = 1/4$) and set the area of each patch to an arbitrary value
of 1 ($\Delta_p = 1$).

@@

We model our pelagic use case loosely on the characteristics of the
Western and Central Pacific Ocean (WCPO) tuna fisheries. We simulate
trajectories of `r n_distinct(names(blue_fauna))` species commonly
caught in the region, including both the highly abundant skipjack tuna
and the heavily depleted oceanic whitetip shark (*Carcharhinus
longimanus*) (@fig-blue-ssb). We use publicly available data on
catch-per-unit-effort of each of these species from the WCPO as a proxy
for baseline habitat distributions, noting that where possible,
fishery-independent abundance indices would be preferable (@fig-blue).

These pelagic species are caught by a longline fleet that primarily
targets large adult tunas such as bigeye and yellowfin for high-grade
consumption, and a purse-seine fishery that primarily targets skipjack
tunas for bulk canning. Selectivities were modeled as logistic for the
longline fleet, and dome-shaped for the purse-seine fleet, with
selectivity values set to broadly match the patterns reported in
relevant [stock
assessments](https://oceanfish.spc.int/en/ofpsection/sam/sam/210-stock-assessment-input-and-result-files-)
for each species. Both fleets operate under open-access dynamics with an
effort cap. The effort cap was set at the level of effort that resulted
in the desired levels of SSB/SSB0 for each species under open-access
dynamics (@fig-blue), intended to simulate a scenario where managers
step in to prevent further expansion of fishing effort in a fully
developed fishery. For forward-simulation, open-access dynamics can
result in effort decreasing in response to profitability, but cannot
result in effort beyond the effort cap set for each fleet.

For this exercise, we focused on using `marlin` to assess resilience of
the selected "rule of thumb" MPA design strategies to a climate-driven
range shift. Specifically, we simulate an extreme example where the
centroid of each population shifts northward at a rate of 0.2 patches
per year, over a 20 year time horizon (@fig-blue-ssb). We designed MPA
networks given the conditions in the starting year, and then held that
network constant over the years of the experiment, running one
simulation with and another without the climate-drive range shift. We
then compared the effects of this range shift on food production and
conservation outcomes from MPA networks designed based on the pre-range
shift world. We ran the simulation on a quarterly level
($\Delta_t = 1/4$) and set the area of each patch to an arbitrary value
of 1 ($\Delta_p = 1$).

@@

# critter_biomass %>%
#   ungroup() %>% 
#   left_join(sci_to_com, by = "critter") %>%
#   select(-critter) %>%
#   rename(critter = common_name) %>%
#   ggplot(aes(step, depletion, color = critter)) +
#   geom_line(linewidth = 1) +
#   scale_y_continuous(limits = c(0, 1), name = "Depletion", expand = expansion(mult = c(0.0,0.05))) +
#   scale_x_continuous(name = "Time Step") +
#   scale_color_discrete(name = "")

@@

((coral_fleet_yield / coral_critter_bio) | (coral_frontier + plot_layout(guides = "keep")) ) & theme(strip.text = element_text(size  = 8), plot.margin = margin(.1, .1, .1, .1, "cm"), panel.spacing = unit(1, "lines"), legend.text = element_text(size = 7))

>>>>

@@

@@

Communities around the world are increasingly looking to spatial strategies for managing marine ecosystems .
For example, the "30x30" movement calls for implementation of marine protected areas (MPAs), a form of spatial management, across 30% of the world's oceans by the year 2030 [@grorud-colvert2021 and references therein].
However , designing spatial management strategies to achieve desired objectives -- which may include recovery and resilience of overfished species, increased food production and economic well being, and equitable distribution of benefits - is not straightforward.
Different species have different resilience to fishing, distributions in space and time, and value to fishing fleets; fishing fleets themselves can have varying ranges of incentives and reactions to policy structures.
These dynamics pose challenges even for single-species assessment and management strategies , which are only amplified when we consider management policies specifically designed around multiple species, fleets, and spatial features in the oceans [@field2006a].

To illustrate, policies such as MPAs designed for both conservation and food production must consider factors such as the optimal size and placement of a protected area network given a wide array of life history, species distributions, exploitation levels, fleet dynamics, and policy-dependent behavior [@reimer2017], all of which may change over time, particularly given the impacts of climate change.
Efforts to effectively design spatial management strategies such as MPAs are further constrained by a lack of empirical evidence describing the size [@ovando2021a] and time required [@nickols2019] for MPAs to produce substantial effects across a range of social-ecological settings [@mcclanahan2021].

Many models have been developed to support the theory and design of spatial management strategies (see @fulton2015 for a review of existing methods).
However, these models have tended to be either highly complex and tactical models designed for use in a specific location, or to be extremely stylized representations intended to provide theoretical insights with less relevance for specific applications.
This lack of accessible models suited for more tactical application presents a challenge to stakeholders charged with marine spatial planning exercises.
It also presents a barrier to the scientific community, wherein comparisons across spatial management simulation studies are clouded by discrepancies in underlying model structures beyond differences in the core phenomenon in question.

To help address this challenge, we present a novel bio-economic modeling tool called [`marlin`](https://danovando.github.io/marlin/) that allows for efficient simulation of spatio-temporal biological and economic dynamics.
`marlin` allows users to simulate the impacts of marine management policies across a range of species targeted by various fleets across heterogeneous and dynamic seascapes.
This model can help users seek Pareto-optimal solutions that produce win-wins or minimize trade-offs across multiple management objectives, such as food production and conservation [@lester2013].

Here we present the core methods and functionality of `marlin`, and demonstrate its use in two case-study applications; a coral reef system and a pelagic system.
Our results show how interacting ecological, economic, and design attributes can affect the degree of benefits, costs, and trade offs between conservation and food production outcomes of MPAs.
More broadly, this paper both demonstrates the functionality of our model and the critical need for considering sufficiently realistic coupled ecological and economic dynamics in policy evaluations.

@@

The front end of the software package accompanying this paper, [`marlin`](https://danovando.github.io/marlin/), is written in R [@rcoreteam2021] to facilitate use, while the underlying population model is written in C++, integrated through the `Rcpp` package [@eddelbuettel2018], for speed.
Users on standard computers should be able to simulate one age-structured population distributed across a 10 by 10 patch system for 20 years in fractions of a second.

@@

We define a few commonly used terms here.
*Yields* refers to the volume of catch from fishing activity.
*Spawning stock biomass* refers to the total biomass of reproductively mature fish in the water, a function of numbers, weight, fecundity, and sexual maturity at age.
We measure the size of the population by th ratio of the *spawning stock biomass* (SSB) in a given time step relative to *unfished spawning stock biomass* (SSB0), SSB/SSB0.
So, a SSB/SSB0 value of 1 means that the population is unfished, a SSB/SSB0 value of 0 means the population is extinct.

@@

These species can be caught by fishing fleets.
A fishing fleet is defined in `marlin` by a set of fishers that have the same fishing skill, prices, and selectivity for individual species (each of which we denote as a *metier*).
For example, both a longline and purse seine fleet may capture bigeye (*Thunnus obesus*) and skipjack tunas, but the longline fleet can be made much more likely to catch larger bigeye than skipjack, and *vice versa*.
Each fishing fleet distributes its fishing effort in space according to a specified spatial allocation rule (see *Spatial Allocation of Effort* ), for example in proportion to profit-per-unit-effort, conditional on management policies in places such as quotas and/or the presence of any spatial restrictions such as MPAs.

@@

`marlin` also allows users to specify closed fishing seasons for one or more species in the system.
When all species in a simulation share the same closed seasons, no fishing occurs in those seasons.
When not all species share the same closed seasons, fishing still occurs in the closed seasons, but the fleet avoids catching the species for which fishing is closed in a given season, assuming no discard mortality.
When closed seasons are in effect, the amount of effort in the next open season is a function of the effort and profits in the last open fishing season.
Lastly, managers can specify locations of no-take MPAs, which can change in size and location if needed.
When MPAs are implemented, fishing effort that used to operate inside the MPAs can either leave the fishery, or be redistributed outside the MPA (the default behavior).

@@

Each fishing fleet decides how to allocate its effort in space based on one of four possible spatial allocation strategies.
The ideal free distribution (IFD) is the standard method for distributing fishing fleets in spaces (see @gillis2003 and references therein).
However, analytical solutions to the IFD present a number of complications for our model.
Our model is capable of tracking multiple fleets at once.
This means that the IFD for a single fleet would commonly be modeled as a Nash-equilibrium based on the actions of each of the individual fleet conditional on the actions of all other fleets.
While possible to implement, this would slow down our model runs to the point of practically preventing large-scale evaluation of the spatial management policies.

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain (CTMC), as described in @thorson2021a.
Within this framework, the model allows for movement to be broken down into three components of *advection* (drifting with currents), *taxis* (active movement towards preferred habitat), and *diffusion* (essentially remaining variation in movement not explained by advection or taxis).
For now, `marlin` focuses just on the diffusion and taxis components of this model, assuming that advection is zero, though future extensions could easily incorporate advection vectors from oceanographic models.
In this way, `marlin` allows users to run anything from a simple Gaussian dispersal kernel up to a system governed by species that passively diffuse out from a core habitat defined by a dynamic thermal range.

We provide a brief overview of the the general CTMC method here, see @thorson2021a for a detailed description.
Under this framework, movement from each patch to each other patch in the system in a given timestep *t* for life stage *a* of species *s* is defined by a movement matrix $\boldsymbol{M}_{t,s,a}$.
$\boldsymbol{M}_{t,s,a}$ is calculated as the matrix exponent of a function of the diffusion $\boldsymbol{D}$ and taxis $\boldsymbol{\tau}$ matrices scaled by the width of the time step $\Delta_{t}$ and the length of the edge of each patch $\Delta_d$ that the model is running under.
So for example, if we are running a model on a monthly timestep, and we define the units of time in years, then $\Delta_{t} = 1/12$.
Similarly, if we are running a model where one patch equals 100km^2^, then $\Delta_d = 10km$.
This parameterization allows users to set the effective area of the spatial domain through two avenues; the number of patches, which effectively scales the resolution of the model, and the area of each patch, which scales the spatial extent of the simulation.

@@

The individual components (*m*) of the movement matrix are filled based on an adjacency matrix, which defines whether two patches are both adjacent and water (as opposed to land or another physical barrier), a diffusion rate $D$ defined in units of area of a patch per unit of time, and a habitat preference function *h* in units of length of a side of a patch per unit time.
For example, if we are defining the time units as years and the distance units as kilometers, for a tuna $D$ might be 1,000 $\frac{KM^2}{Year}$.
We then use parameters $\Delta_{t}$ and $\Delta_{d}$ parameters to translate the diffusion rate $D$ to match the time step and patch size used in a simulation.
This "scale free" parameterization means that appropriate value of $D$ can be identified for a species and then set, regardless of the time step or patch size used in the simulation model itself.
The taxis component of the movement process is a function of the difference in habitat quality *h*.
The habitat preference function itself can take any form the user wishes.
Exponentiating the difference in the habitat preference function between patches turns the taxis matrix into a multiplier of the diffusion rate *D*.
As such, when creating habitat layers for simulation, users can tune the scale of the habitat gradient function to result in realistic multipliers of the diffusion rate.
This parameterization ensure that the off-diagonal elements of the movement matrix $\boldsymbol{M}_{t,s,a}$ are all non-negative, a requirement of the CTMC method.

$$ 
m_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}^2}De^{\frac{\Delta_t(h(p2,t,s,a) - h(p1,t,s,a))}{\Delta_d}} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} m_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
\end{cases}
$$ {#eq-diffusion}

@@

While this CTMC approach to movement simulation is to date not commonly seen in the marine modeling literature, it has numerous advantages that warrant its seeming complexity.
First, when only diffusion is present, the model will generalize to the familiar dynamics of a Gaussian dispersal kernel.
However, the time component *t* in @eq-movement allows the movement model to be run on arbitrarily fine time steps.
Second, the taxis model allow for more active habitat choices by species, allowing us to simulate preferences of species in space and time efficiently.
Lastly, this form has the advantage that the parameters are directly estimable from real data.
So, if provided with for example spatial abundance data and a tagging study, users can estimate the diffusion and taxis movement parameters in the same manner as @thorson2021a, and then pass their estimated parameters to `marlin` for simulation (so long as the estimating method uses the same functional form as the movement model in `marlin`).

@@

Recruitment (i.e. the number of age 1 individuals entering the population) follows Beverton-Holt dynamics parameterized around steepness (*h*) with log-normally distributed recruitment deviates $\epsilon$.
When steepness is one recruitment is independent of spawning biomass.
As steepness approaches 0.2 recruitment becomes a linear function of spawning biomass.
We allow for five variants in the timing of density dependent recruitment, building off of @babcock2011 :

@@

4.  Post-dispersal density dependence: Larvae are distributed throughout the system, and then density dependent recruitment occurs based on the density of spawning biomass at the destination patch.

@@

Fisheries management is often concerned with measuring stock status relative to maximum sustainable yield (MSY) based reference points, though the exact level of stock status relative to MSY reference points desired by management or societal objectives may vary widely.
MSY based reference points present a problem for a multi-fleet and spatial-temporal model such as `marlin`.
MSY and the fishing mortality rate that would produce MSY, $F_{MSY}$, are a function of fishery selectivity.
Fishery selectivity in this model can vary by fleet, and species can be distributed unevenly in both space and time.
This means that the net effective fishing selectivity on a species can vary depending on the dynamics at a given moment, making the definition of an equilibrium concept such as MSY challenging.

@@

Each of the case studies contains too many parameters and options to be succinctly presented in the text here, and we are concerned that attempting to duplicate all of the parameters of the model within the text would simply introduce errors and confusion.
As such, and since these results are for illustrative purposes only, readers should consult the accompanying code to view the precise details of each simulation.
Targeted tactical applications must carefully consider and document all decisions made around model parameters.

@@

We make use of three potential "rule of thumb" MPA design strategies:

@@

3.  *Spawning Aggregation*: MPAs are centered on the location of a known spawning aggregation

To our knowledge, MPAs are never randomly placed; their locations are decided based on stakeholder negotiation given available information and objectives.
In theory, the design of MPA networks can be optimized through the use of a modeling framework, and depending on the validity of the model, this process is likely to produce better outcomes than manually-designed strategies [@rassweiler2012; @rassweiler2014].
However, designing optimal MPA networks becomes increasingly difficult as the range of objectives and the complexity of the model increase.
Therefore, while `marlin` provides some support for [optimization](https://danovando.github.io/marlin/articles/optimize-mpas.html), we focus here on the design and performance of these more "rule of thumb" design strategies that may be more accessible to a wider range of users.
We allow all MPAs to be designed in a "mosaic" fashion in these examples, but users can easily extend the analysis to compare outcomes between "contiguous" (MPA is made up of one continuously connected block) and "mosaic" (MPAs can be separated in space) MPA designs [@pons2022] by developing and passing their own MPA design algorithms to the model.

#### Coastal Coral Reef Fishery

In our coastal coral reef example, we model the dynamics of four tropical Pacific species: a grouper (*Epinephelus fuscoguttatus*), a shallow-reef snapper (*Lutjanus malabaricus*), a deep-water snapper (*Pristipomoides filamentosus*), and a reef shark (*Carcharhinus amblyrhynchos*).
The simulated groupers undergo a mass migration to a spawning aggregation once per year, followed by the sharks.
Shallow-reef snappers stay in reefs closer to shore above a steep drop-off year-round, while deep-water snappers stay in the deeper reefs past the drop-off (@fig-coral).

These species are targeted by two different fleets.
Fleet One primarily targets the grouper and near-shore snapper populations, but will land any incidentally captured sharks.
Fleet One has a logistic selectivity pattern for all species, as they retain any fish caught for consumption or sale.
both grouper and shallow-reef snapper due to demand for "plate sized" fish , and logistic selectivity for sharks, as any shark caught is retained.
Fleet One is totally dependent on fishing for their livelihood, meaning the local community takes advantage of every possible opportunity to fish, and as such we model it as a "constant effort" fishery.
Do to having less efficient boats, Fleet One has a higher cost per distance coefficient than Fleet Two.
Fleet One's home port is located on the site of the grouper spawning aggregation.

Fleet Two is a more commercial fleet that primarily targets the deep-water snapper population.
This fleet primarily sells their catch to local restaurants and distributors where plate-sized fish are prized, and so for both snapper and grouper Fleet Two has a dome-shaped selectivity pattern [@kindsvater2017].
While plate-sized deep snapper are the primary target of Fleet Two, we model Fleet Two's selectivity for deep snapper as logistic due to high levels of discard mortality for deep-water snapper resulting from barotrauma.
Fleet Two catches groupers, though less than Fleet One, and receives no price for sharks due to the requirements of a certification program through which they sell their deep-water snapper.
Accidental captures (bycatch) of sharks do occur, which results in mortality.
Fleet Two operates under open-access dynamics, as fishing is not the only means of subsistence for this community; short-term effort expands and contracts in response to profitability of the primarily grouper-driven fishery.
Fleet two's home port is located further north from Fleet One.

We use `marlin` to simulate the outcomes for both food production and conservation for each of the species and both of the fleets as a function of both MPA size and MPA design strategy.
For this exercise, MPAs are placed with perfect information and have no design constraints for continuity (i.e. the selected MPA design can be as patchwork as dictated by the data).
We ran the simulation on a quarterly level ($\Delta_t = 1/4$) and set the area of each patch to be 5km^2^ ($\Delta_p = \sqrt{5}$), using 400 patches for a total area of 2,000 KM^2^.

@@

We model our pelagic use case loosely on the characteristics of the Western and Central Pacific Ocean (WCPO) tuna fisheries.
We simulate trajectories of `r n_distinct(names(blue_fauna))` species commonly caught in the region, including both the highly abundant skipjack tuna and the heavily depleted oceanic whitetip shark (*Carcharhinus longimanus*) (@fig-blue-ssb).
We use publicly available data on catch-per-unit-effort of each of these species from the WCPO as a very rough proxy for baseline habitat distributions, noting that where possible, fishery-independent abundance indices would be preferable (@fig-blue).

These pelagic species are caught by a longline fleet that primarily targets large adult tunas such as bigeye and yellowfin for high-grade consumption, and a purse seine fishery that primarily targets skipjack tunas for bulk canning.
Selectivities were modeled as logistic for the longline fleet, and dome-shaped for the purse seine fleet, with selectivity values set to broadly match the patterns reported in relevant [stock assessments](https://oceanfish.spc.int/en/ofpsection/sam/sam/210-stock-assessment-input-and-result-files-) for each species.
Both fleets operate under open-access dynamics with an effort cap.
The effort cap was set at the level of effort that resulted in the desired levels of SSB/SSB0 for each species under open-access dynamics (@fig-blue), intended to simulate a scenario where managers step in to prevent further expansion of fishing effort in a fully developed fishery.
For forward-simulation, open-access dynamics can result in effort decreasing in response to profitability, but cannot result in effort beyond the effort cap set for each fleet.

For this exercise, we focused on using `marlin` to assess resilience of the selected Target Fishing and Rate "rule of thumb" MPA design strategies to a climate-driven range shift.
Specifically, we simulate an extreme example where the centroid of each population shifts northward at a rate of 0.33 patches per year, over a 20 year time horizon (@fig-blue-ssb).
We designed MPA networks given the conditions in the starting year, and then held that network constant over the years of the experiment, running one simulation with and another without the climate-drive range shift.
We then compared the effects of this range shift on food production and conservation outcomes from MPA networks designed based on the pre-range shift world.
We ran the pelagic simulation at a quarterly level ($\Delta_t$ = 1/4 year), and set the area of each cell to be roughly 97,000 KM^2^ across 144 patches each with a side length of roughly 311 KM, for a total area of 14,000,000 KM^2^, broadly commensurate with the area of the WCPO.

@@

@@

((coral_fleet_yield / coral_critter_bio) | (coral_frontier + plot_layout(guides = "keep")) ) & theme(strip.text = element_text(size  = 6), plot.margin = margin(.1, .1, .1, .1, "cm"), panel.spacing = unit(0.5, "lines"), legend.text = element_text(size = 6))
